TBD

### 2018-12-19 paper/paper.tex
updated paper

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style  which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.2
\documentclass[modern]{aastex62}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
Gaussian processes are frequently used to model variability in astrophysical time-series. In addition to studying variability in stars and AGN, Gaussian 
process regression can be used to account for correlated noise in light curves of exoplanetary transits and micro-lensing events. One of the chief limitations of this 
technique is the computational time required to compute a Gaussian process model for large data-sets. This also limits prospects for using Gaussian processes with 
multi-dimensional datasets, although sparse-matrix approximations can make GP regression on multi-dimensional datasets feasible for some applications. Here we 
present a method based on the \celerite GP implementation which enables fast computation of two-dimensional gaussian process noise models for datasets with a 
small second dimension, such as multi-band photometry. This method enables the use of GP regression with multi-wavelength transit and micro-lensing light curves, 
and may have applications to radial velocity and direct imaging data as well.

\end{abstract}

@@

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below. 
\tableofcontents

\section{introduction}\label{sec:intro}
	Gaussian processes have been used extensively in astrophysics to model stochastic variability across space and time. 
	They have been used to study variable stars and AGN, to model the extragalactic distribution of dust, and to model correlated noise in 
	photometric and radial velocity observations of exoplanet host stars.	\gordontodo{X-ray binaries?}

	\gordontodo{The following paragraph might make more sense in the exomoons section.}
	Gaussian process noise models are most advantageous when correlated noise dominates over white noise. 
	This is already the case for state-of-the-art radial velocity observations, and transit 
	photometry is not far behind. Future space-based observatories such as James Webb 
	will enable high-precision photometry in which stellar variability is the main barrier to detecting ever more shallow transit signals. 
	Stellar variability which obscures transit signal can be a result of spot and faculae crossings, asteroseismic oscillations, or variable 
	granulation on the star's surface. GPs are able to \gordontodo{something about how a GP model is non-parametric and the advantages thereof? Also agnostic 
	to the source of variability (seems related)}

	\gordontodo{This is where we come in...}

	\gordontodo{Are GPs applicable to instrumental noise as well? Seems best to account for instrumental effects with models that account for known factors since 
	they might not be well-described by a gaussian process, right?}

	\gordontodo{What about other methods of accounting for astrophysical noise? i.e. cosine filtering, polynomials, etc. --- also related to GP as non-parametric 
	noise model}

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}{\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood function given by
	\begin{equation}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} = \ln\ p(\bvec{y}|X, \bvec{\theta}, \bvec{\alpha}) = 
			-\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi)
	\end{equation}
	\gordontodo{I'm using $X$ here because DFM does, but wouldn't $\bvec{x}$ be more clear?}
	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative of the log-likelihood), one can obtain an estimate of the hyperparameters $\bvec{\theta}$ and 
	$\bvec{\alpha}$. Furthermore, Markov Chain Monte Carlo (MCMC) methods can be used to sample the likelihood in hyperparameter-space. \gordontodo{expand 
	on this.}
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they are computationally expensive, with computation of the likelihood function scaling as 
		$\mathcal{O}(N^3)$ due to the matrix inversion. This is especially problematic for use cases such as MCMC which require calling the likelihood function many 
		times. This has prevented the adoption of Gaussian process noise models for datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice. The \celerite method overcomes this issue by reducing the time to compute the likelihood function to $\mathcal{O}(NJ^2)$ for one 
		class of kernel functions; where J is the number of \celerite terms making up the kernel function of the GP. This class of kernel functions are given by 
		\begin{equation}
			k_\alpha(t_n, t_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the white noise, and $\tau_{nm} = |t_n-t_m|$. 
		The coefficients $a$ and $c$ may be complex.

		For kernel functions of this form, the covariance matrix can be written as a semi-separable matrix of rank $2J$:
		\begin{equation}
			K = A + \mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)
		\end{equation}
		where $U$ and $V$ have entries given by: 
		\begin{eqnarray}
			U_{n, 2j-1} &=& a_je^{-c_jt_n}\cos(d_jt_n) + b_je^{-c_jt_n}\sin(d_jt_n) \nonumber \\
			U_{n, 2j} &=& a_je^{-c_jt_n}\sin(d_jt_n) - b_je^{-c_jt_n}\cos(d_jt_n) \nonumber \\
			V_{m, 2j-1} &=& e^{c_jt_m}\cos(d_jt_m) \nonumber \\
			V_{m, 2j} &=& e^{c_jt_m}\sin(d_jt_m)
		\end{eqnarray}
		and A is a diagonal matrix:
		\begin{equation}
			A_{n,n} = \sigma_n^2 + \sum_{j=1}^Ja_j
		\end{equation}
		Cholesky decomposition can be performed on this matrix in $\mathcal{O}(NJ^2)$ time, as described in the appendix.

		\gordontodo{Add discussion of the versatility of this kernel, and it's connection to stochastically driven harmonic oscillators.}

		\gordontodo{Add discussion of semiseperability --- the reader should understand that the limitation of \celerite is that only kernels of this form can be 
		represented by a semi-seperable matrix, which is what makes moving to two dimensions slightly more complicated than ordinary.}

\section{2d gaussian processes}
	\subsection{kronecker structure of the 2d covariance matrix}
	In two dimensions the kernel function depends on two coordinates. We will continue to use
	\begin{equation}
		\bvec{t} = \expandvec{t_1}{t_N}
	\end{equation}
	to refer to the first coordinate, and introduce
	\begin{equation}
		\bvec{u} = \expandvec{u_1}{u_M}
	\end{equation}
	to refer to the second coordinate. In most cases, we will demand that $M << N$. We then define $\bvec{x}$ to be the list of points
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1), (t_1, u_2), \cdots (t_1, u_M)}{(t_N, u_M)}
	\end{equation}
	We can also write this as
	\begin{equation}
		x_i = (t_{\lfloor(i-1)/M\rfloor + 1}, u_{i - \lfloor(i-1)/M\rfloor M})
	\end{equation} 
	The covariance between $x_i$ and $x_j$ is then given by
	\begin{equation}
		k_{\bvec{\alpha}\bvec{\beta}}(x_i, x_j) = \sigma_i^2\delta_{ij} + q_\bvec{\alpha}(\tau_{ij})\circ r_\bvec{\beta}(\nu_{ij})
	\end{equation}
	where $\tau_{ij} = |t_{\lfloor(i-1)/M\rfloor + 1}-t_{\lfloor(j-1)/M\rfloor + 1}|$, 
	$\nu_{ij} = |u_{i - \lfloor(i-1)/M\rfloor M}-u_{j - \lfloor(j-1)/M\rfloor M}|$; 
	$q$ and $r$ are one-dimensional kernel functions which depend on the hyperparameters $\bvec{\alpha}$ and 
	$\bvec{\beta}$ respectively; 
	and $\circ$ is some binary operation. 
	If $Q_\bvec{\alpha}$ and $R_\bvec{\alpha}$ are the covariance matrices corresponding to 
	$q_\bvec{\alpha}$ and $r_\bvec{\alpha}$ such that
	\begin{equation}
		\left[Q_\bvec{\alpha}\right]_{ij} = q_\bvec{\alpha}(\tau_{ij})\ \ \ \mathrm{and}\ \ \ \left[R_\bvec{\beta}\right]_{ij} = r_\bvec{\beta}(\tau_{ij})
	\end{equation} 
	then, using $q_{ij}$ and $r_{ij}$ as shorthand for the entries in $Q_\bvec{\alpha}$ and $R_\bvec{\beta}$, $K_\bvec{\alpha\beta}$ can be represented as: 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + 
			\begin{bmatrix}
				q_{11}\circ R_\bvec{\beta} & ... & q_{1N}\circ R_\bvec{\beta} \\
				\vdots & \ddots & \\
				q_{N1}\circ R_\bvec{\beta} & ... & q_{NN}\circ R_\bvec{\beta}
			\end{bmatrix}
	\end{equation}
	where $[\Sigma]_{ij} = \sigma_i^2\delta_{ij}$ and
	\begin{equation}
		q_{pq}\circ R_\bvec{\alpha} = 
			\begin{bmatrix}
				q_{pq}\circ r_{11} & ... & q_{pq}\circ r_{1M} \\
				\vdots & \ddots & \\
				q_{pq}\circ r_{M1} & ... & q_{pq}\circ r_{MM}
			\end{bmatrix}
	\end{equation}
	If the operator $\circ$ represents multiplication, we have 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + Q_\bvec{\alpha} \otimes  R_\bvec{\beta}
	\end{equation}
	where $\otimes$ is the Kronecker product. If the covariance in $\bvec{t}$ is described by a \celerite kernel can write 
	\begin{eqnarray}
		K &=& \Sigma + \left[\mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)\right]\otimes R \\
		&=& \Sigma + \mathrm{tril}\left[ (U\otimes R)(V\otimes I_M)^\T\right] + \mathrm{triu}\left[ (V\otimes I_M)(U\otimes R)^\T\right]
	\end{eqnarray}
	This covariance has the same semi-separable form as in the one-dimensional case, but now $K$ is an $(NM\times NM)$ matrix of rank $2JM$. 
	This means that we can apply the same Cholesky decomposition algorithm (described in the appendix) as in the 1d case to compute the GP in 
	$\mathcal{O}(NJ^2M^3)$. While our implementation only handles the Kronecker product case, 
	his result holds in the case that $\circ$ is an operation other than multiplication, so long as $\circ$ has the property
	\begin{equation}
		(a\circ B)(c\circ D) = (ac)\circ(BD)
	\end{equation}
	for scalars $a$ and $b$ and matrices $B$ and $D$. 
	%where $a\circ B$ is the matrix 
	%\begin{equation}
	%	a\circ B = \begin{bmatrix}
	%		a\circ b_{11} & a\circ b_{12} & \cdots \\
	%		a\circ b_{21} &  a\circ b_{22} & \\
	%		\vdots & & \ddots & 
	%	\end{bmatrix}
	%\end{equation}
	\gordontodo{See appendix for proof}

	\subsection{sampling the distribution}
	A gaussian process can be used to simulate correlated noise. If $L$ is the lower triangular factor from the Cholesky decomposition of the 
	$(N\times N)$ covariance matrix $K$, and $\bvec{z}$ is a vector of length $N$ containing random numbers drawn for a standard normal distribution, i.e.
	\begin{equation}
		\bvec{z} \sim \mathcal{N}(1, 0)
	\end{equation}
	then $\bvec{x} = \bvec{\mu_x} + L\bvec{z}$ is a vector with entries correlated by $K$ and with mean $\bvec{\mu_x}$, i.e.
	\begin{equation}
		\bvec{x} \sim \mathcal{N}(\bvec{\mu_x}, K)
	\end{equation}
	\gordontodo{fix this so that $\mu$ depends on $\theta$ and is a function of $\bvec{t}$ or something like that. Compuational cost should be $\mathcal{O}(NJM^2)$}

	\subsection{interpolation and extrapolation}	
	\gordontodo{Bit more complicated --- if both dimensions are celerite processes then the scaling is better than in the case where $R$ is not celerite. You have notes on this.}

\section{example: exomoons}
	\gordontodo{What if I used Kepler-1625 as an example? Could simulate WFC3 GRISM data and show that moon could be validated with Celerite2D?}

>>>>

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style  which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.2
\documentclass[preprint2]{aastex62}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
Gaussian processes are a model of stochastic variability frequently used in analyzing astrophysical time-series. 
They have been used to study variability in light curves of stars and AGN, as well as the spatial 
distribution of the extragalactic dust distribution. In the realm of stellar variability, GP regression 
is used to characterise stellar rotation periods, search for transit signals in noisy data, and analyze 
RV curves. One of the chief limitations of this method is the computational time incurred in computing 
a GP model for large or multi-dimensional data-sets. While approximate methods can make GP regression 
feasible for some of these applications, in others exact methods are preferable. Here 
we present a method based on the \celerite GP implementation which enables fast, exact 
computation of two-dimensional GP models for data-sets with a small second dimension. This includes 
multi-bandpass photometry of transit and microlensing light curves. Our method also may have applications 
to RV and direct imaging methods. 
\end{abstract}

@@

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below. 
%\tableofcontents
\section{introduction}\label{sec:intro}
    A Gaussian process (GP) is a generalization of a Gaussian distribution to 
    function-space \citep{Rasmussen2006}. GPs have been used extensively in 
    astrophysics as a model of stochastic variability across both space and time, 
    to study a diverse range of phenomena including variable stars, AGN\citep{Kelly2014}, the 
    extragalactic dust distribution\citep{Sale2018}, and gravitational wave signals\citep{Moore2016}. 
    They have also been used to account 
    for correlated noise in photometric and RV measurements of 
    exoplanet hosts\citep{Aigrain2016,Jones2017,Rajpaul2015}.

    GP models offer advantages over other methods of modeling stochastic variability 
    because they are able to account for correlated variations without the imposition 
    of a parametric model. \gordontodo{more about advantages of GPs}

    The primary limitation of GP methods is that most GP operations are 
    computationally expensive. For instance, the common tasks 
    of computing likelihoods and sampling from the GP require 
    $\mathcal{O}(N^3)$ operations for $N$ data points. This is 
    especially problematic for use cases which require repeated calls 
    to the likelihood function, such as minimization and performing 
    MCMC. This has prevented the adoption of GP methods for data-sets 
    larger than $N\sim 10^4$ points\citep{Deisenroth2015}.

    The issue of computational expense can be partially overcome either by approximating 
    the GP or by restricting the user to a subset GPs for which the computation 
    can be sped up\citep{Rasmussen2006}. In the latter category we 
    find \celerite\citep{Foreman-Mackey2017}, a fast one-dimensional GP 
    method which increases the speed of most common GP computations to $\mathcal{O}(NJ^2)$ 
    where $J$ is typically small compared to $N$. \celerite achieves this speedup by 
    limiting the user to GPs with a specific functional form of the covariance. 
    Another limitation of \celerite is that the method is only applicable to 
    one-dimensional GPs. Here we present a method to rectify this difficiency of 
    the original \celerite method. Our extension of \celerite computes two-dimenional 
    GPs on an $N$ by $M$ grid in $\mathcal{O}(NJ^2M^3)$ operations. This is an improvement 
    over the general case in which the computation of the same GP would require $\mathcal{O}(N^3M^3)$ 
    operations. Because the speed of our computation is still sharply dependent on $M$, it 
    is most useful in cases where $M << N$. An example would be modeling variability in 
    multi-bandpass photometry where $M$ would represent the number of bands.

    In this paper we first review the \celerite method before introducing our extension 
    of the original method. We then give a detailed example of GP regression on multi-band 
    transit photometry to recover transit parameters in the presence of significant 
    correlated noise. Finally, we discuss other potential applications to our method and 
    conclude with an outline of future work in which we hope to further reduce the 
    computational expense of a two-dimensional GP. This would enable the use of 
    GP methods on even larger data-sets and potentially make feasible GP computations 
    in higher dimensions.

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}
		    {\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. 
	The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to 
	as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood 
	function given by
	\begin{eqnarray}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} &=& \ln\ 
		    p(\bvec{y}|\bvec{x}, \bvec{\theta}, \bvec{\alpha}) \\ 
		&=& -\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi) \nonumber
	\end{eqnarray}
	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative 
	of the log-likelihood), one can obtain an estimate of the hyperparameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$. Markov Chain Monte Carlo (MCMC) methods can be used to 
	sample the posterior distribution over the hyperparameters. 
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they 
		are computationally expensive, with computation of the likelihood function 
		scaling as $\mathcal{O}(N^3)$ due to the matrix inversion. This is especially 
		problematic for use cases such as MCMC which require calling the likelihood 
		function many times. This has prevented the adoption of Gaussian process noise models for 
		datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice. The \celerite method overcomes this limitation by reducing 
		the time to execute most GP computations to linear time for one class of 
		kernel functions\citep{Foreman-Mackey2017}. We give a brief description of the \celerite method 
		and its limitations below.

		Consider a one-dimensional gaussian process evaluated at the coordinates
		\begin{equation}
			\bvec{x} = \expandvec{t_1}{t_N}
		\end{equation}
		The \celerite kernel is given by
		\begin{equation}
			k_\alpha(t_n, t_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the gaussian-distributed white noise, and $\tau_{nm} = |t_n-t_m|$. 
		The coefficients $a$ and $c$ may be complex, in which case we introduce $b$ and $d$ which are the imaginary parts of $a$ and $c$ respectively. 
		This kernel defines a \celerite model with $J$ terms.

		For a kernel function of this form, the covariance matrix is a symmetric, semiseparable with semiseparability rank $R=2J$. A matrix of this type can 
		be written in terms of two generator matrices $U$ and $V$, both of size $(N\times 2J)$, and a diagonal matrix $A$:
		\begin{equation}
		\label{eqn:KUV}
			K = A + \mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)
		\end{equation}	
		In the case of our covariance matrix, the generator matrices are specified by:	
		\begin{eqnarray}
			U_{n, 2j-1} &=& a_je^{-c_jt_n}\cos(d_jt_n) + b_je^{-c_jt_n}\sin(d_jt_n) \nonumber \\
			U_{n, 2j} &=& a_je^{-c_jt_n}\sin(d_jt_n) - b_je^{-c_jt_n}\cos(d_jt_n) \nonumber \\
			V_{m, 2j-1} &=& e^{c_jt_m}\cos(d_jt_m) \nonumber \\
			V_{m, 2j} &=& e^{c_jt_m}\sin(d_jt_m)
		\end{eqnarray}
		and A is given by:
		\begin{equation}
			A_{n,n} = \sigma_n^2 + \sum_{j=1}^Ja_j
		\end{equation}
		Computing the Cholesky decomposition for this covariance matrix can be accomplished in $\mathcal{O}(NJ^2)$ operations, allowing for the 
		fast computation of the GP likelihood function\footnote{The algorithm is described in appendex \ref{sec:celerite_algorithm}}.

		%\gordontodo{Add discussion of the versatility of this kernel, and it's connection to stochastically driven harmonic oscillators.}
		%
		%\gordontodo{Add discussion of semiseperability --- the reader should understand that the limitation of \celerite is that only kernels of this form can be 
		%represented by a semi-seperable matrix, which is what makes moving to two dimensions slightly more complicated than ordinary.}
		The main limitation of \celerite is that the method only applies to one class of kernels. However, the \celerite kernel overcomes this limitation 
		by being both versatile in its ability to approximate other kernels, and also directly applicable to many problems. The versatility of this 
		kernel is demonstrated in \gordontodo{figure reference} which shows approximations to several popular kernels achieved by appropriate choices 
		of the \celerite coefficients $a_j$ and $c_j$.

		One model of stellar oscillations models the star as a stochastically-driven damped harmonic oscillator\citep{Anderson1990}. The solution to the equations describing 
		such an oscillator is a GP with a \celerite kernel with coefficients 
		\begin{eqnarray}
			a_{j\pm} &=& \frac{1}{2}S_0\omega_0\left[1\pm\frac{1}{\sqrt{1-4Q^2}}\right] \\
			c _{j\pm}&=& \frac{\omega_0}{2Q}\left[1 \mp \sqrt{1-4Q^2}\right]
		\end{eqnarray}
		where $S_0$ is proportional to the power at the un-damped frequency of the oscillator, $\omega_0$, and $Q$ is a parameter known as 
		the quality factor of the oscillator. Variations in the stellar radius due to these oscillations gives rise to brightness variations in the star's 
		light curve \gordontodo{reference?}. 
		This means that a \celerite GP can be used as a model for stellar variability due 
		to p-mode oscillations. Additionally, setting $Q=1/\sqrt{2}$ results in a PSD 
		which has been used to describe granulation-driven stellar variability\citep{Kallinger2014}.

\section{2d gaussian processes}
	In addition to being limited to one class of kernels \celerite, as formulated above, is inherently one-dimensional. For many applications, such 
	as modeling noise in single-bandpass light curves or RV time series GP computations in one dimension are all that is necessary. However in 
	other cases such as modeling noise in simultaneous multi-bandpass light curves or disentangling noise in RV data using multiple activity 
	indicators, two-dimensional GP methods are necessary. We outline a method to compute a GP in two dimensions utilizing the \celerite 
	method for the larger of the two dimensions to speedup most GP operations by a factor of $\sim\mathcal{O}(N^2)$ over ordinary 2D GPs.

	In two-dimensions the independent variable is
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1)}{(t_N, u_N)}
	\end{equation}
	If the covariance between two points satisfies 
	\begin{equation}
	\label{eqn:multiplication_condition}
		k_\alpha(x_n, x_m) = q_\alpha(t_n, t_m)r_\alpha(u_n, u_m)
	\end{equation}
	then the covariance matrix has the structure of a Hadamard matrix product (element-wise 
	multiplication )
	\begin{equation}
		\left[K_\alpha\right]_{nm} = \left[Q_\alpha\right]_{nm}\left[R_\alpha\right]_{nm}
	\end{equation}
	or 
	\begin{equation}
		K = Q\circ R
	\end{equation}
	Now consider a GP evaluated on a $(N\times M)$ grid. We begin by defining
	\begin{eqnarray}
		\bvec{t'} &=& \expandvec{t'_1}{t'_N} \\
		\bvec{u'} &=& \expandvec{u'_1}{u'_M}
	\end{eqnarray}
	to be vectors containing only the unique $t$ and $u$ values in increasing order. 
	%, such that 
	%\begin{equation}
	%	\bvec{x} = \expandvec{((t'_1, u'_1)\cdots(t'_N,u'_1), (t'_1, u'_2)\cdots(t'_N,u'_2)}{(t'_1, u'_M)\cdots(t'_N, u'_M)}
	%\end{equation}
	We also define the corresponding covariance matrices for each dimension: 
	\begin{subequations}
	\label{eqn:QR_defs}
	\begin{eqnarray}
		\left[Q'_\alpha\right]_{nm} &=& q_\alpha(t'_n, t'_m) \\
		\left[R'_\alpha\right]_{pq} &=& r_\alpha(u'_p, u'_q) 
	\end{eqnarray}
	\end{subequations}
	In this case, assuming (\ref{eqn:multiplication_condition}), the full covariance matrix can be written as a Kronecker product between these 
	two one-dimensional covariances\footnote{See appendix \ref{sec:matrix_prod} for a proof of (\ref{eqn:kronecker_K}) and notes on the Hadamard and Kronecker products. }: 
	\begin{equation}
	\label{eqn:kronecker_K}
		K_\alpha = Q'_\alpha \otimes R'_\alpha
	\end{equation}
	Since the rest of this discussion will focus on the case of gridded data, we will drop the prime notation and use $Q_\alpha$ and $R_\alpha$ to refer to the 
	matrices in (\ref{eqn:QR_defs}). In the following discussion we assume that the covariance in $t$ is described by a \celerite kernel, and that 
	the covariance in $u$ is arbitrary. This means that $Q_\alpha$ can be expressed in terms of the generator matrices $U$ and $V$.  Following equation 
	(\ref{eqn:KUV}), in the absence of a white noise term, we can write the covariance matrix as follows\footnote{See appendix \ref{sec:Kproof} for proof}:
	\begin{eqnarray}
	\label{eqn:KUVR}
		%\nonumber K &=& A\otimes R + \mathrm{tril}(UV^\T)\otimes R + \mathrm{triu}(VU^\T)\otimes R \\
		%\nonumber &=& A\otimes R + \mathrm{tril}((UV^\T)\otimes (RI_M)) + \mathrm{triu}((VU^\T)\otimes (RI_M) ) \eqnum{\value{equation}} \\
		\nonumber K &=& \left(A + \mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)\right)\otimes R \\
		&=& A' + \mathrm{tril}(U'V'^\T) + \mathrm{triu}(V'U'^T)
	\end{eqnarray}
	where
	\begin{eqnarray}
		A' &=& A\otimes R \\
		U' &=& U\otimes R \\
		V' &=& V\otimes I_M
	\end{eqnarray}
	To include white noise we add a white noise term to each element of A:
	\begin{equation}
		A'_{n, n} \to A'_{n, n} + \sigma_n^2
	\end{equation}
	From the form of equation \ref{eqn:KUVR} we can see that the covariance is, as before, semiseparable. However Kronecker 
	multiplication by $R$ and $I_M$ has increased the size of the generator matrices from $(N\times 2J)$ to $(NM\times 2JM)$. As a result,  
	the number of operations necessary to compute the Cholesky decomposition to increase from $\mathcal{O}(NJ^2)$ to $\mathcal{O}(NJ^2M^3)$. 
	The number of operations for sampling from and extrapolating/interpolating with the two-dimensional GP scales the same. Some improvement 
	can be made in the scaling for extrapolation and interpolation if the covariance in the second dimension can be expressed via 
	a \celerite kernel. In this case the algorithm requires $\mathcal{O}(nNP + mMP)$ operations where $P$ is the number of points at which the prediction is evaluated and $n, m$ are constants. The 
	Cholesky composition algorithm as well as algorithms for interpolation/extrapolation and sampling from the GP are described in appendix 
	\ref{sec:algorithms}.

	It should be noted that the scaling of our method is still relatively poor as we add points to our grid in the second dimension. This 
	is why the method is most suitable for cases where the second dimension is much smaller than the first, i.e. where $M << N$. 
	This condition applies to the important cases of modeling noise in multi-bandpass observations. 
	%Into appendix? 
	%So that $Q'_\alpha$ is a symmetric $(N\times N)$ matrix (with hyperparameters $\alpha$) 
	%describing the covariance for a one-dimensional GP evaluated along $\bvec{t}$ and 
	%$R'_\alpha$ is the same for $\bvec{u}$.  These matrices can be related back to the un-primed $Q_\alpha$ and $R_\alpha$ 
	%matrices via the Kronecker product.
